Weighted Visibly Pushdown Automata and Automated Music Transcription

### 2021-04-28 main.tex
def. SWVPA

\newcommand{\weight}{\mathsf{weight}}
\newcommand{\init}{\mathsf{in}}
\newcommand{\final}{\mathsf{out}}

@@

%\noindent %properties:
A semiring $\Semiring$ is \emph{commutative} if $\otimes$ is commutative.
%\francesco{How we deal with that since it is non commutative?}
It \emph{bounded}~\cite{Mohri02semiring} if $\forall x \in \dom(\Semiring), \one \oplus x = \one$,
and \emph{idempotent} if for all $x \in \Semiring$, $x \oplus x = x$.
%We restrict in the following of \emph{idempotent} semirings.
Note that every bounded semiring is idempotent:
by boundedness, 
$\one \oplus \one = \one$, and idempotency follows by multiplying
both sides by $x$ and distributing. %

@@

\section{SW Visibly Pushdown Automata}
We follow the approach of~\cite{Mohri03EDWA} for the computation of distances...

\subsection{SW Automata and Transducers}
The following definition of weighted transducers over infinite alphabets 
generalizes weighted transducers over finite alphabets, see  e.g.~\cite{Mohri03EDWA}, 
by considering weight functions generalizing the guards of symbolic automata

@@

\begin{definition}
A weighted transducer $T$ over the input and output alphabet $\Sigma$ and $\Delta$ 
and the semiring $\Semiring$ is a tuple
$T = \< Q, \init, \weight, \final >$,
where $Q$ is a finite set of states, 
$\mathsf{in} : Q \to \Semiring$, 
respectively $\mathsf{out} : Q \to \Semiring,$
is a function defining the weight for entering, 
respectively leaving, a state, 
and $\weight$ is a transition %weight 
function of $Q \times Q$ into %$\Phi$.
$\< \Phi_0, \Phi_\Sigma, \Phi_\Delta, \Phi_{\Sigma, \Delta}>$.
\end{definition}
We extend the above transition function into a function from
$Q \times (\Sigma \cup \{ \epsilon \}) \times (\Delta \cup \{ \epsilon \}) \times Q$
into $\Semiring$, also called $\weight$ for simplicity, such that 
for all $q, q' \in Q$, $a \in \Sigma$, $b \in \Delta$, 
and with 
$\< \phi_\epsilon, \phi_\Sigma, \phi_\Delta, \phi_{\Sigma, \Delta}> = \weight(q, q')$, 
\[
\begin{array}{rcl}
\weight(q, \epsilon, \epsilon, q') & = & \phi_\epsilon\\
\weight(q, a, \epsilon, q') & = & \phi_\Sigma(a)\\
\weight(q, \epsilon, b, q') & = & \phi_\Delta(b)\\
\weight(q, a, b, q') & = & \phi_{\Sigma, \Delta}(a, b)
\end{array}      
\]
%More precisely, $Q \times Q$, 
%resp. $Q \times \Sigma \times Q$,
%$Q \times \Delta \times Q$,
%$Q \times \Sigma \times \Delta \times Q$,
%into resp. $\Phi_0$, $\Phi_\Sigma$, $\Phi_\Delta$ $\Phi_{\Sigma, \Delta}$.

@@

\subsection{Distance between words or languages}
distance $d$:
defined over $\Sigma^* \times \Sigma^*$ 
into a semiring  $\Semiring = ( \Semiring, \oplus, \zero, \otimes, \one)$.

\paragraph{Edit-Distance.}
...algebraic definition of edit-distance of Mohri, in \cite{Mohri03EDWA}
% Mehryar Mohri 
% Edit-distance of weighted automata: General definitions and algorithms
% International Journal of Foundations of Computer Science 14.06 (2003): 957-982.
%
\noindent
Let $\Omega = \Sigma \cup \{ \epsilon \} \times \Sigma \cup \{ \epsilon \} \setminus \{ (\epsilon, \epsilon) \}$,
and let $h$ be the morphism from $\Omega^*$ into $\Sigma^* \times \Sigma^*$  
defined over the concatenation of strings of $\Sigma^*$ (that removes the $\epsilon$'s).
%
\noindent
An \emph{alignment} between 2 strings  $s, t \in \Sigma^*$ is an element $\omega \in \Omega^*$ 
such that $h(\omega) = (s, t)$.
%
\noindent
We assume a base cost function $\Omega$ : $\delta: \Omega \to S$, extended to $\Omega^*$ as follows  
(for $\omega \in \Omega^*$): 
\(
\displaystyle\delta(\omega) = \bigotimes_{0 \leq i < |\omega|} \delta(\omega_i)
\).

\subsection{SW Visibly Pushdown Automata}

@@

>>>>

\def\SWT{\textsf{SWT}\xspace}
\def\SWA{\textsf{SWA}\xspace}
\def\SWVPA{\textsf{SWVPA}\xspace}
\def\weight{\mathsf{weight}}
\def\wei{\mathsf{w}}
\def\init{\mathsf{in}}
\def\final{\mathsf{out}}
\newcommand{\call}[1]{\ensuremath \langle_{#1}}
\newcommand{\return}[1]{\ensuremath {}_{#1}{\rangle}} % $\prescript{}{a}{)}$
\def\Sigmai{{\Sigma_\mathsf{i}}}
\def\Sigmac{{\Sigma_\mathsf{c}}}
\def\Sigmar{{\Sigma_\mathsf{r}}}
\def\Phii{{\Phi_\mathsf{i}}}
\def\Phic{{\Phi_\mathsf{c}}}
\def\Phir{{\Phi_\mathsf{r}}}
\def\Phicr{{\Phi_\mathsf{cr}}}
\def\weii{{\wei_\mathsf{i}}}
\def\weic{{\wei_\mathsf{c}}}
\def\weir{{\wei_\mathsf{r}}}
\def\weie{{\wei_\mathsf{e}}}
\newcommand{\config}[2]{\ensuremath \genfrac{[}{]}{0pt}{}{#1}{#2}}

@@

and let $\< \Semiring, \oplus, \zero, \otimes, \one>$ be a {semiring},

@@

\section{SW Automata and Transducers}
\label{section:transducer}
We follow the approach of~\cite{Mohri03EDWA} for the computation of distances
between words with transducers.

The following definition of weighted transducers over infinite alphabets 
generalizes weighted transducers over finite alphabets, see  e.g.~\cite{Mohri03EDWA}, 
by considering weight functions generalizing the guards of symbolic automata

@@

%(SWT)
\begin{definition}
A \emph{symbolic-weighted transducer} $T$ over the input and output alphabet $\Sigma$ and $\Delta$ 
and the semiring $\Semiring$ is a tuple
$T = \< Q, \init, \wei, \final >$,
where $Q$ is a finite set of states, 
$\mathsf{in} : Q \to \Semiring$, 
respectively $\mathsf{out} : Q \to \Semiring,$
are functions defining the weight for entering, 
respectively leaving, a state, 
and $\wei$ is a transition %weight 
function from $Q \times Q$ into %$\Phi$.
$\< \Phi_0, \Phi_\Sigma, \Phi_\Delta, \Phi_{\Sigma, \Delta}>$.
\end{definition}
%
We extend the above transition function into a function from
$Q \times (\Sigma \cup \{ \epsilon \}) \times (\Delta \cup \{ \epsilon \}) \times Q$
into $\Semiring$, also called $\wei$ for simplicity, such that 
for all $q, q' \in Q$, $a \in \Sigma$, $b \in \Delta$, 
and with 
$\< \phi_\epsilon, \phi_\Sigma, \phi_\Delta, \phi_{\Sigma, \Delta}> = \wei(q, q')$, 
\[
\begin{array}{rcl}
\wei(q, \epsilon, \epsilon, q') & = & \phi_\epsilon\\
\wei(q, a, \epsilon, q') & = & \phi_\Sigma(a)\\
\wei(q, \epsilon, b, q') & = & \phi_\Delta(b)\\
\wei(q, a, b, q') & = & \phi_{\Sigma, \Delta}(a, b)
\end{array}      
\]
%More precisely, $Q \times Q$, 
%resp. $Q \times \Sigma \times Q$,
%$Q \times \Delta \times Q$,
%$Q \times \Sigma \times \Delta \times Q$,
%into resp. $\Phi_0$, $\Phi_\Sigma$, $\Phi_\Delta$ $\Phi_{\Sigma, \Delta}$.

@@

The symbolic-weighted transducer $T$ defines a mapping 
from the pairs of strings of $\Sigma^* \times \Delta^*$ 
into the weights of~$\Semiring$,
based on the following intermediate function $\weight_T$
defined recursively for every $q, q' \in Q$, 
%$a \in \Sigma$, $b \in \Delta$ 
for every strings of $s \in \Sigma^*$, $t \in \Delta^*$:
\[
\begin{array}{rccl}
\weight_T(q, s, t, q') & = & & \wei(q, \epsilon, \epsilon, q')\\
 & & \oplus \displaystyle\bigoplus_{\begin{array}{c}
                                      \scriptstyle q'' \in Q\\[-2pt]
                                      \scriptstyle s = au, a \in \Sigma
                                      \end{array}} &
     \wei(q, a, \epsilon, q'') \otimes \weight_T(q'', u, t, q')\\
 & & \oplus \displaystyle\bigoplus_{\begin{array}{c}
                                    \scriptstyle q'' \in Q\\[-2pt]
                                    \scriptstyle t = bv, b \in \Delta
                                    \end{array}} &
      \wei(q, \epsilon, b, q'') \otimes \weight_T(q'', s, v, q')\\ 
      & & \oplus \displaystyle\bigoplus_{\begin{array}{c}
                                         \scriptstyle q'' \in Q\\[-2pt]
                                         \scriptstyle s = au, a \in \Sigma\\[-2pt]
                                         \scriptstyle t = bv, b \in \Delta
                                         \end{array}} &
      \wei(q, a, b, q'') \otimes \weight_\A(q'', u, v, q')\\ 
\end{array}
\]
Recall that by convention, an empty sum with $\oplus$ is $\zero$. 
%
\noindent
The weight associated by $T$ to  $\< s, t> \in \Sigma^* \times \Delta^*$
is then defined as follows: 
\[
T(s, t)  = 
\displaystyle\bigoplus_{q, q' \in Q} \mathsf{in}(q) 
\mathop{\otimes} \weight_T(q, s, t, q') \mathop{\otimes} \mathsf{out}(q').
\]

A \emph{symbolic weighted automata} (\SWA) $A = \< Q, \init, \weight, \final >$
over $\Sigma$ and $\Semiring$ 
is defined in a similar way by simply omitting the output symbols,
\ie $\wei$ is a function of~$Q \times Q$ into %$\Phi$.
$\< \Phi_0, \Phi_\Sigma >$, 
or equivalently from~$Q \times (\Sigma \cup \{ \epsilon \}) \times Q$ into~$\Semiring$.

\begin{proposition}
Given a \SWT $T$ 
over $\Sigma$, $\Delta$ and $\Semiring$, 
and a word $s \in \Sigma^*$, 
one can construct a \SWA 
$A_{s, T}$ such that for all $t \in \Delta^*$, 
$A_{s, T}(t) = T(s, t)$.
\end{proposition}
The construction time and size of $A_{s, T}$ are $O(| s | . \| T \|)$.

\section{SW Visibly Pushdown Automata}
The following model generalizes Symbolic VPA~\cite{DAntonyAlur14SVPDA}
from Boolean semirings to arbitrary semiring weight domains.

Let $\Sigma$ be an input alphabet, finite (large) or infinite,
that we assume partitioned into :
\begin{itemize}
\item a set $\Sigmai$ of \emph{internal symbols} denoted $a$,
\item a set $\Sigmac$ of \emph{call symbols} denoted $\call{a}$,
\item a set $\Sigmar$ of \emph{return symbols} denoted $\return{a}$.
\end{itemize}
In order to simplify notations, and following the definition of Section~\ref{section:transducer}, 
we shall write respectively~$\Phii$, $\Phic$, $\Phir$ and~$\Phicr$
for~$\Phi_\Sigmai$, $\Phi_\Sigmac$, $\Phi_\Sigmar$ and~$\Phi_{\Sigmac, \Sigmar}$,

\begin{definition}
A \emph{Symbolic Weighted Visibly Pushdown Automata} (\SWVPA) $A$ 
over the input $\Sigma = \Sigmai \uplus \Sigmac \uplus \Sigmar$
and the semiring $\Semiring$ is a tuple
$T = \< Q, P, \init, \weii, \weic, \weir, \weie, \final >$,
where $Q$ is a finite set of states, 
$P$ is a finite set of stack symbols, 
$\mathsf{in} : Q \to \Semiring$, 
respectively $\mathsf{out} : Q \to \Semiring,$
are functions defining the weight for entering, 
respectively leaving, a state, 
and 
$\weii : Q \times Q \to \Phii$,  
$\weic : Q \times Q \times P \to \Phic$,  
$\weir : Q \times P \times Q \to \Phicr$,  
$\weie : Q \times Q \to \Phir$,  
are transition functions.
\end{definition}
%
Similarly as in Section~\ref{section:transducer}, 
we extend the above transition functions as follows
for all $q, q' \in Q$, $p \in P$, 
$a \in \Sigmai$, 
$\call{c} \in \Sigmac$, 
$\return{r} \in \Sigmar$, 
overloading the names for simplicity:
\[
\begin{array}{lll}
\weii: Q \times \Sigmai \times Q \to \Semiring & 
\weii(q, a, q') = \phi_\mathsf{i}(a) & 
\mathrm{where~} \phi_\mathsf{i} = \weii(q, q')\\
%
\weic: Q \times \Sigmac \times Q \times P \to \Semiring & 
\weic(q, \call{c}, q', p) = \phi_\mathsf{c}(\call{c}) & 
\mathrm{where~} \phi_\mathsf{c} = \weic(q, q', p)\\
%
\weir: Q \times \Sigmac \times P \times \Sigmar \times Q \to \Semiring & 
\weir(q, {\call{c}},  p, {\return{r}}, q') = \phi_\mathsf{r}({\call{c}},  {\return{r}}) & 
\mathrm{where~} \phi_\mathsf{r} = \weir(q, p, q')\\
%
\weie: Q \times \Sigmar \times Q \to \Semiring & 
\weie(q, {\return{r}}, q') = \phi_\mathsf{e}({\return{r}}) &
\mathrm{where~} \phi_\mathsf{e} = \weie(q, q')\\
\end{array}      
\]

\noindent
The intuition is the following for the above transitions.
\begin{description}
\item $\weii$ : read the input internal symbol $a$, change state to $q'$.
\item $\weic$ : read the input symbol $\call{c}$, push it to the stack along with $p$, change state to $q'$.
\item $\weir$ : when the stack is not empty, 
      read and pop from stack a pair made of $\call{c}$ and $p$, 
      read the input symbol $\return{r}$, change state to $q'$.
      In this case, the weight function $\phi_\mathsf{r}$ checks a matching between the call and return symbols.
\item $\weie$ : when the stack is empty, 
      read the input symbol $\call{r}$, change state to $q'$.
\end{description}

We give now a formal definition of these transitions of the automaton $A$
in term of a weight value
computed by an intermediate function $\weight_A$.
In the case of a pushdown automaton, a configuration is composed 
of a state $q \in Q$ and a stack content $\theta \in \Theta^*$, where $\Theta = \Sigmac \times P$.
Therefore, $\weight_A$ is a function from 
$Q \times \Theta^* \times \Sigma^* \times Q \times \Theta^*$ into $\Semiring$.
\[
\begin{array}{rcl}
\weight_A\bigl(\config{q}{\theta}, a u, \config{q'}{\theta'}\bigr) & = & 
 {\displaystyle\bigoplus_{q'' \in Q}} \weii(q, a, q'') 
 \otimes \weight_A\bigl(\config{q''}{\theta}, u, \config{q'}{\theta'}\bigr)\\
%
\weight_A\bigl(\config{q}{\theta}, {\call{c}} u, \config{q'}{\theta'}\bigr) & = & 
 {\displaystyle\bigoplus_{\begin{array}{c}
                          \scriptstyle q'' \in Q\\[-2pt]
                          \scriptstyle p \in P
                          \end{array}}}
 \weic\bigl(q, {\call{c}}, q'', p\bigr) 
 \otimes \weight_A\bigl(\config{q''}{{\call{c}}\, p\cdot \theta}, u, \config{q'}{\theta'}\bigr)\\[1mm]
%
\weight_A\bigl(\config{q}{{\call{c}}\, p\cdot \theta}, {\return{r}} u, \config{q'}{\theta'}\bigr) & = & 
 {\displaystyle\bigoplus_{q'' \in Q}} 
 \weir\bigl(q, {\call{c}}, p, {\return{r}}, q''\bigr) 
 \otimes \weight_A\bigl(\config{q''}{\theta}, u, \config{q'}{\theta'}\bigr)\\
%
\weight_A\bigl(\config{q}{\bot}, {\return{r}} u, \config{q'}{\theta'}\bigr) & = & 
 {\displaystyle\bigoplus_{q'' \in Q}} \weie(q, {\return{r}}, q'') 
 \otimes \weight_A\bigl(\config{q''}{\bot}, u, \config{q'}{\theta'}\bigr)\\
\end{array}
\]
where $\bot$ denotes the empty stack and ${\call{c}}\, p\cdot \theta$ 
denotes a stack with the pair made of ${\call{c}}$ and $p$ on its top and $\theta$
as the rest of stack.

\noindent
The weight associated by $A$ to $s \in \Sigma^*$
is then defined as follows by empty stack computation: 
\[
A(s)  = 
{\displaystyle\bigoplus_{q, q' \in Q}} \textstyle
\mathsf{in}(q) \mathop{\otimes} 
\weight_A\bigl(\config{q}{\bot}, s, \config{q'}{\bot}\bigr) 
\mathop{\otimes} \mathsf{out}(q').
\]

% The weight of a transition acts as a guard: 
% a transition is activated for a symbol $`a`$ iff its weight $`\phi(a) \neq 0`$ (or $`\psi(a, b) \neq 0`$), 
% the absorbing element.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% APPENDIX                                                                   %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\appendix

\section{Edit-Distance}

%\subsection{Distance between words or languages}
...algebraic definition of edit-distance of Mohri, in \cite{Mohri03EDWA}
% Mehryar Mohri 
% Edit-distance of weighted automata: General definitions and algorithms
% International Journal of Foundations of Computer Science 14.06 (2003): 957-982.
distance $d$ over $\Sigma^* \times \Sigma^*$ 
into a semiring  $\Semiring = ( \Semiring, \oplus, \zero, \otimes, \one)$.

%\noindent
Let $\Omega = \Sigma \cup \{ \epsilon \} \times \Sigma \cup \{ \epsilon \} \setminus \{ (\epsilon, \epsilon) \}$,
and let $h$ be the morphism from $\Omega^*$ into $\Sigma^* \times \Sigma^*$  
defined over the concatenation of strings of $\Sigma^*$ (that removes the $\epsilon$'s).
%
\noindent
An \emph{alignment} between 2 strings  $s, t \in \Sigma^*$ is an element $\omega \in \Omega^*$ 
such that $h(\omega) = (s, t)$.
%
\noindent
We assume a base cost function $\Omega$ : $\delta: \Omega \to S$, extended to $\Omega^*$ as follows  
(for $\omega \in \Omega^*$): 
\(
\displaystyle\delta(\omega) = \bigotimes_{0 \leq i < |\omega|} \delta(\omega_i)
\).
